---
title: "BDCD Business Analytics Lab exercise"
author: "Prof Di Cook, with Steph Kobakian, Stuart Lee, Mitch O'Hara-Wild"
date: "Econometrics & Bus Stat, Monash, Clayton campus, 5/4/2018"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE)
```

## Motivation

Business analytics involves mathematics and computing. This lab exercise has a little of both. For the coding part, you will want to copy, pull apart and put together again. 

Sports provide a lot of statistics about players and teams. In this exercise we are going to look at the statistics collected for the Women's Australian Rules Football leagues over the past two years. We have stats for players and for teams. These are averages across all the games: Kicks, Handballs, Dispatch efficiency (%), Marks, Frees_Agst, Goals, Behinds, Goal_assists, Time_On_Ground. 

Materials for the workshop can be downloaded from [https://github.com/BDCD18/BusAn](https://github.com/BDCD18/BusAn).

Data collected from [http://www.afl.com.au/womens/matches/stats](http://www.afl.com.au/womens/matches/stats).

## Exercise 1: Background work

Find the web site for the competition, to determine the answers for these questions.

1. What team won the competition in 2017, and 2018? Who were the runners-up in each year?
2017 Winners: Adelaide Crows
2017 Runner Ups: Western Bulldogs
2018 Winners: Western Bulldogs
2018: Runner ups: Brisbane Lions


2. What players were awarded best and fairest in each year?
2017 B&F: Erin Phillips (Adelaide Crows)
2018 B&F: Emma Kearney (Western Bulldogs)


```{r fig.margin=TRUE, fig.width=3.5, fig.height=7}
library(tidyverse)
players <- read_csv("AFLW/data/players.csv")
ggplot(players, aes(x=Kicks, y=Disp_eff)) + 
  geom_point() + 
  facet_wrap(~Year, ncol=1) + 
  theme(aspect.ratio=1) +
  xlab("Av Kicks/Game") +
  ylab("% Dispatch Efficiency")
```

## Exercise 2: Explore the data

1. We have written a web app, using R, that you can access by 
- open the R project "AFLW"
- open the file "app.R"
- Click "Run App"
2. 
    a. Which player had the highest average kicks in 2017? 2018?
  Highest average kicks:
2017: Erin Phillips 
2018: Emma Kearney 

    b. Is the player who has the most kicks, also good at dispatching the ball?
    Player also good at dispatching?
(2017) Erin Phillips: No. Dispatch Efficiency: 57.5%
(2018) Emma Kearney: No. Dispatch Efficiency: 60.5%

b(i) Name a few of the players who managed to dispatch 100% of the time, and also the ones who never manage to dispatch the ball.
Names of players who dispatched with: 
2017: 
100% efficiency: Kim Mickle, Natalie Plane
0% efficiency: Lou Wotton, Taryn Preistly 
2018:
100% efficiency: Sarah Lampard
0% efficiency: Ruby Blair, Jessica Anderson, Louise Stephenson 


    c. What was the highest average number of goals by a player in both seasons? Who achieved this?
    Highest average goal 2017: 2 (Darcy Vescio)
Highest average goal 2018: 1.6 (Jess Wuetschner)

     c(i)Is this different from the player with the most kicks? How?
     The players who had the most kicks were not the players with the highest average amount of goals, this could be explained by a players different role in a team (defense, full forward, half forward, etc). Erin and Emma (highest average of kicks) may not be goal shooters and may be in defense thus explaining the difference from the players with the most kicks with the players with the most goals. 



```{r fig.margin=TRUE, fig.width=3.5, fig.height=3.5}
players <- players %>% mutate(GWS = ifelse(Club == "GWS", "yes", "no"))
players_sel <- players %>% filter(Year == "2018") %>%
  filter(Player %in% c("Alicia Eva", "Jessica Dal Pos"))
ggplot(filter(players, Year == "2018"), aes(x=Kicks, y=Disp_eff, colour=GWS)) + 
  geom_point() + 
  theme(aspect.ratio=1, legend.position = "none") +
  xlab("Av Kicks/Game") +
  ylab("% Dispatch Efficiency") + 
  geom_label(data=players_sel, aes(label = Player), 
             vjust="top", hjust="right", alpha=0.6) + 
  scale_colour_brewer(palette="Dark2")
```

## Exercise 3: Checking the news

1. The news article [AFL's edict to women players: make the game more entertaining](https://www.theaustralian.com.au/sport/afl/afls-edict-to-women-players-make-the-game-more-entertaining/news-story/44e9b896759ca76edf166b38499d149d) suggests that the organisers felt the games were too low scoring in the first season. 
    a. Has this changed in season 2? 
    
    The low scoring situation has changed,
According to the charts of the comparison between goals and kicks in 2017 and 2018, we can clearly see dots between the part of 0.5 to 1.5 has a significant increase which means the goals grew in season 2018. 

    b.  Make a plot of goals vs kicks to help to answer this question.

2. [AFLW best and fairest: AFL rejects claim that votes were awarded to wrong player](https://www.foxsports.com.au/afl/womens-afl/aflw-best-and-fairest-afl-rejects-claim-that-votes-were-awarded-to-wrong-player/news-story/c0ae1205ccfecc7481dcf4edc45ed692) 
    a. What two players are mentioned in the article?
    Jessica Dal Pos & Alicia Eva  (teammates at Greater Western Sydney)

    b. Colour the team discussed in the article.
    
    c. Does the data suggest that if one of these players got votes, it was a mistake? And that they should have gone to the other player? If so, why?
    In 2018 Alicia Eva had an average of 17.8 disposals per game, while Jessica Dal Pos averaged 11 disposals. This gap of 6.8 disposals per game on average is a large difference and supports the fact that Jessica received votes over her teammate alicia. However, when comparing the two players dispatch efficiency, the statistics show that Jessica averaged a 63.6% efficiency which is larger than Alicia's average dispatch efficiency of 56%. This statistic supports the fact that the votes could have rightfully been awarded to Jessica over Alicia.


```{r fig.margin=TRUE, fig.width=3.5, fig.height=7}
players_sub17 <- players %>%
       filter(Year == "2017") 
players_sub18 <- players %>%
       filter(Year == "2018") 
players_sub_mat17 <- as.matrix(players_sub17[,4:12])
players_sub_mat17 <- apply(players_sub_mat17, 2, scale)
players_sub_mat18 <- as.matrix(players_sub18[,4:12])
players_sub_mat18 <- apply(players_sub_mat18, 2, scale)
players_mds17 <- cmdscale(dist(players_sub_mat17), k=2)
players_mds18 <- cmdscale(dist(players_sub_mat18), k=2)
players_mds_df17 <- as_tibble(players_mds17)
players_mds_df17$Player <- players$Player[players$Year == "2017"]
players_mds_df17$Year <- "2017"
players_mds_df18 <- as_tibble(players_mds18)
players_mds_df18$Player <- players$Player[players$Year == "2018"]
players_mds_df18$Year <- "2018"
players_mds_df <- bind_rows(players_mds_df17, players_mds_df18)
p1 <- players_mds_df %>% filter(Year == "2017", Player == "Erin Phillips")
p2 <- players_mds_df %>% filter(Year == "2018", Player == "Emma Kearney")
pl <- bind_rows(p1, p2)
ggplot(players_mds_df, aes(x=V1, y=V2, label=Player)) + 
  geom_point() + facet_wrap(~Year, scales="free", ncol=1) +
  geom_point(data=pl, colour = "orange", size=2) +
  geom_label(data=pl, aes(label = Player), 
             vjust="top", hjust="left", alpha=0.6)
```

## Exercise 4: Similarity of players

- Change the variables being combined to make the low-dimensional picture. Does the plot change? Explain why this happens based on the formulae given in the lecture notes.
The plot graph changes. If we add more and more variables, compute the difference between all variables of two player, we can examine the difference between them. And two different player in some area will be similar (such as Kim Mickle and Natalie Plane in Kicks, Hand balls and Dispatch Efficiency, but not so close for all variables)

- Color the points by teams. Is there much difference between teams? If so, or if not, what does this mean?
In terms of all round players within each team, the 2017 data plot suggests that players are similar is all round skill as most dots are clustered in the centre of the graph. Interestingly, potential outliers, who are further away from the cluster graph but closer to each other, are present in the two winning teams from WB (runners up for 2017) and ADEL (premiership winners for 2017). WB’s and ADEL’s more scattered data distribution furthermore suggests that the two teams are more similar in terms of all round players than other teams in 2017. However, in 2018 some WB’s players are more scattered compared to ADEL indicating a difference between the two teams. The proximity of dots belonging to Adelaide Crows is small meaning that most players in Adelaide crows are similar. Additionally, apart from high performing players in Western Bulldogs, the remaining players are similar to Adelaide Crow players meaning that the a majority of Adelaide Crow players and Western Bulldog players are similar in all round skills. 
- Find the most valuable player awardee in each year, and look at the nearest neighbours of those players. Were they also nominees for the award?
AFLW Players' MVP
2017: Winner  Erin Phillips (204)
She is similar to Ellie Blackburn but she not nominees for award in 2017
2018: Courtney Gum 
Nearest players that are not nominees are:
- Monique Conti(in handballs and kicks)
-Stephanie Chiocci (in handballs and kicks)
-Brianna Davey (in handballs and kicks)
-Meg Hutchins (in frees and behinds)

```{r fig.margin=TRUE, fig.width=3.5, fig.height=7}
library(forcats)
ggplot(players, aes(x=fct_relevel(Club, c("COLL", "CARL", "ADEL", "MELB", "WB", "BL", "FRE", "GWS")), y=Kicks)) + 
  geom_boxplot() +
  facet_wrap(~Year, ncol=1) + xlab("Club")
```

## Exercise 5: Your turn to code

1. Easy tasks: 
    a. Change the title from "Exploring the AFLW statistics" to "Women's AFL Statistics Exploration"
    b. Change the label of tab 1 from "Data" to "Basic plots"
2. Medium task: 
    a. Change the tab 2 title to "Boxplots"
    b. Remove the "X" variable input menu
    c. Change the type of plot on tab 1 to be a side-by-side boxplot of the statistic by Club. 
3. Difficult: Make a new app to study the similarity of the teams. The steps to do this are:
    a. Copy and paste the entire directory code for the "AFLW" app. Rename it. 
    b. Read in the teams data, and pass this into the analysis functions.  
    c. Change quite a few things: the list of variables to use for the MDS, name of the tab, label by Club instead of Player, .... 

With your new app, using ONLY this set of statistics Kicks, Handballs, Dispatch efficiency, Frees for and against, Goals and Behinds, which teams are most similar based on being close together in the MDS plot? Which are very different from each other?

## Turn in

Each group needs to provide to the instructor:

1. A document with answers to each of the questions
2. A copy of your app code
3. A presentation slide with a plot of the data, and a sentence about what you've learned about women's AFL. (HINT: This part will have the most points for judging the best team, so be creative and informative.)

